Re-examining cross-cultural similarity judgments using language statistics

### 2023-05-09 code/4_paper/manuscript/manuscript.tex
cogsci ready

Alternatively, these judgments could be shaped by cross-cultural
differences in the \emph{input} to similarity judgment, that is, the
statistics of the linguistic and/or physical environment as manifest in
everyday experiences. Perhaps when faced with the triad task,
participants from all cultures follow the same process for
conceptualizing similarity, but rely on language or culture-specific
input to this process. If we observe a difference in categorization
between East Asian and Western participants, it could be that members of
both groups use the same procedure (considering similarity that is
influenced by both taxonomic and thematic relations), but that the input
to this procedure differ between cultures, with East Asian participants
exposed to more support for thematic similarity in their experience in
comparison to their Western counterparts.

@@

We measured taxonomic versus thematic similarity matching in a
forced-choice word triad task in three populations. Following Ji et al.
(2004), we measured preferences in the US and China. In addition, we
collected data in a novel context: Vietnam. Vietnam is a Southeast Asian
country that borders China and has historically been greatly influenced
by Chinese culture (Hui, 2002). Therefore, it serves as a suitable
cultural context to investigate whether the claim made by Ji et al.
(2004) and previous studies -- that Eastern and Western cultures have
different notions of similarity -- extends beyond mainland China. In
addition to these replication and extension questions, we tested whether
fastText vectors from corpora corresponding to each language context
(English, Mandarin, and Vietnamese) are good predictors for similarity
judgments in each population.

@@

We excluded 8 US participants, 16 CN participants, and 62 VN
participants, who missed two or more of the 11 attention check
questions.

In doing so, we deviated from our preregistered exclusion criterion,
which used a more stringent policy, excluding participants who missed
any of the 10 attention checks. This approach would have substantially
reduced our sample size, disproportionately affecting the VN sample
(resulting in a final sample of 109 US, 132 CN, and 57 VN participants).
We report results with the less stringent exclusion criterion, but we
performed analyses with both samples and found broadly similar results.
Any differences are noted in the Results \& Discussion section.

@@

To give an intuition for our model, consider again the cow:
grass/chicken triad. We retrieved word vectors for ``cow'' and
``grass'', and calculated the cosine distance between these vectors. We
did the same for ``cow'' and ``chicken.'' Our similarity prediction is
inversely proportional to the ratio of cosine distance of these pairs.
This is because a larger cosine distance means the word vectors are
further apart, and thus the words are less similar. For example, if the
cosine distance of thematic cow-grass is 0.7 and the cosine distance of
taxonomic cow-chicken is 0.3 (the more similar of the two), then our
model predicts, correspondingly, that 30\% of responses to the triad
will be grass, and the other 70\% chicken.

@@

We observed the strongest preference for thematic matching in China (M =
0.65, SD = 0.11), followed by Vietnam (M = 0.6, SD = 0.11), and then the
US (M = 0.56, SD = 0.17; see Figure 1).

@@

\caption[Fixed effect sizes of each corpus lexical statistics (cosine distance proportion) when included as a predictor for China, Vietnam, and US responding, respectively]{Fixed effect sizes of each corpus lexical statistics (cosine distance proportion) when included as a predictor for China, Vietnam, and US responding, respectively. The English corpus is the best predictor for US response, and the Mandarin corpus is the best predictor for China response.}\label{fig:coeffs_cos}
\end{figure}
\end{CodeChunk}

However, in all cultural contexts, adding the other two corpora produced
a significantly better fit than the identical model with only the
corresponding corpus included as a predictor (US responses:
\(\chi^2\)(2) = 7.93, p = 0.019; VN responses: \(\chi^2\)(2) = 13.72, p
= 0.001; CN responses: \(\chi^2\)(2) = 10.56, p = 0.005). This analysis
suggests that culture-specific input to similarity judgments (as proxied
by language statistics) do not fully explain cross-cultural differences
in similarity judgment, whether due to noise or sparsity in the
estimates of language statistics or true additional variance picked up
by embeddings from other corpora.

@@

In contrast, we found some support for the environmental statistics
account: word embeddings based on language-specific corpora were a good
predictor for the corresponding country's similarity judgments, even
when other corpus statistics were included, and even for filler triads
without a thematic/taxonomic structure. Overall, our results provide
evidence that cross-cultural differences in similarity judgments are
related to patterns in linguistic statistics (which also vary across
cultures).

@@

There are still many open questions for this account, however. Our
operationalization of environmental statistics with word embeddings
likely captures both language inputs and culture-specific
conceptualization of similarity. Further investigation of the semantic
spaces captured by cross-cultural word embedding models is required.
Future work should also aim to provide a more specific computational
account of how lexical co-occurrence might guide categorization
preference beyond the simple proportion-of-similarity model tested here,
and investigate other person- or triad-specific factors that might
influence how environmental statistics affect similarity judgments.

@@

@@

@@

@@

@@

>>>>

Alternatively, these judgments could be shaped by cross-cultural
differences in the \emph{input} to similarity judgment, that is, the
statistics of the linguistic and/or physical environment as manifest in
everyday experiences. Perhaps when faced with the triad task,
participants from all cultures follow the same process for
conceptualizing similarity, but rely on language or culture-specific
input to this process. If we observe a difference in categorization
between East Asian and Western participants, it could be that members of
both groups use the same procedure (considering similarity that is
influenced by both taxonomic and thematic relations), but that the input
to this procedure differ between cultures, with East Asian participants
exposed to more support for thematic similarity in their experience in
comparison to their Western counterparts. \textbf{\emph{Because this is
a question about naturally-occurring differences in real-world reasoning
contexts, it is difficult to address experimentally. In a controlled
study, we could manipulate participants' input and determine whether
differing inputs influence similarity judgments (e.g. McDonald \&
Ramscar, 2001), but this would not address the actual sources of
cross-cultural variation in similarity judgments. Here, we draw on
real-world data to estimate variation across cultures.}}

@@

We measured taxonomic versus thematic similarity matching in a
forced-choice word triad task in three populations. Following Ji et al.
(2004), we measured preferences in the US and China. In addition, we
collected data in a novel context: Vietnam. Vietnam is a Southeast Asian
country that borders China and has historically been greatly influenced
by Chinese culture (Hui, 2002). \textbf{\emph{We were unable to find
cultural indices specifically on thematic/taxonomic reasoning, but
Vietnam shows greater similarity to China than the US across the
Hofstede dimensions of cultural comparison (2018) and is often cited for
its similarity to more canonical exemplars of East Asian culture
(Hirschman \& Loi, 1996; Whitmore, 1984; Woodside, 1998), for example,
patterning with East Asian countries rather than Southeast Asian
countries on a measure of family values (Minkov \& Hofstede, 2012).}}
Therefore, Vietnam serves as a suitable cultural context to investigate
whether the claim made by Ji et al. (2004) and previous studies -- that
Eastern and Western cultures have different notions of similarity --
extends beyond mainland China. In addition to these replication and
extension questions, we tested whether fastText vectors from corpora
corresponding to each language context (English, Mandarin, and
Vietnamese) are good predictors for similarity judgments in each
population.

@@

\textbf{\emph{We excluded 8 US participants (4\%), 16 CN participants
(8\%), and 62 VN participants (31.16\%\footnote{The high rate of
  exclusion within our Vietnamese sample may reflect some unfamiliarity
  with attention checks, which could have produced exploratory
  responding to these questions. However, because exploratory responding
  to attention checks is indistinguishable from chance responding (from
  bots or inattentive participants), we did not analyze this data
  further.}), who missed three or more of the 10 attention check
questions.}}

In doing so, we deviated from our preregistered exclusion criterion,
which used a more stringent policy, excluding participants who missed
any of the 10 attention checks. This approach would have substantially
reduced our sample size, disproportionately affecting the VN sample
(resulting in a final sample of 109 US, 132 CN, and 57 VN participants).
We report results with the less stringent exclusion criterion, but we
performed analyses with both samples and found broadly similar results.
Any differences are noted \textbf{\emph{with a footnote}} in the Results
\& Discussion section.

@@

{\centering \includegraphics{figs/method-1}

\caption[Thematic response prediction based on cosine distance of the triad cow]{Thematic response prediction based on cosine distance of the triad cow: grass/chicken.}\label{fig:method}
\end{figure}
\end{CodeChunk}

To give an intuition for our model, consider again the cow:
grass/chicken triad. We retrieved word vectors for ``cow'' and
``grass'', and calculated the cosine distance between these vectors. We
did the same for ``cow'' and ``chicken.'' Our similarity prediction is
inversely proportional to the ratio of cosine distance of these pairs
(Figure 1). This is because a larger cosine distance means the word
vectors are further apart, and thus the words are less similar. For
example, if the cosine distance of thematic cow-grass is 0.7 and the
cosine distance of taxonomic cow-chicken is 0.3 (the more similar of the
two), then our model predicts, correspondingly, that 30\% of responses
to the triad will be grass, and the other 70\% chicken.

@@

We observed the strongest preference for thematic matching in China (M =
0.65, SD = 0.11), followed by Vietnam (M = 0.6, SD = 0.11), and then the
US (M = 0.56, SD = 0.17; see Figure 2).

@@

\caption[Fixed effect sizes of each corpus lexical statistics (cosine distance proportion) when included as a predictor for China, Vietnam, and US responding, respectively]{Fixed effect sizes of each corpus lexical statistics (cosine distance proportion) when included as a predictor for China, Vietnam, and US responding, respectively. Asterisks indicate corpora that are significant predictors. The English corpus is the best predictor for US response, and the Mandarin corpus is the best predictor for China response.}\label{fig:coeffs_cos}
\end{figure}
\end{CodeChunk}

However, in all cultural contexts, adding the other two corpora produced
a significantly better fit than the identical model with only the
corresponding corpus included as a predictor (US responses:
\(\chi^2\)(2) = 7.93, p = 0.019; VN responses: \(\chi^2\)(2) = 13.72, p
= 0.001; CN responses: \(\chi^2\)(2) = 10.56, p = 0.005).
\textbf{\emph{This analysis suggests that culture-specific input to
similarity judgments (as proxied by language statistics) do not fully
explain cross-cultural differences in similarity judgment. This could be
due to noise or sparsity in the estimates of language statistics, so
that adding additional corpora simply improves fit by expanding the
effective size of the model corpus. Alternatively, it could be due to
true additional cultural-specific variance picked up by embeddings from
other corpora --- this is a question for future work.}}

@@

In contrast, we found some support for the environmental statistics
account: word embeddings based on language-specific corpora were a good
predictor for the corresponding country's similarity judgments, even
when other corpus statistics were included, and even for filler triads
without a thematic/taxonomic structure. \textbf{\emph{This evidence is
correlational and cannot evaluate causes of variation in similarity
reasoning; indeed, it may be that cultural variation in similarity
reasoning shapes the linguistic variation we measure, but we note that
this account does not explain the fit to filler triads.}} Overall, our
results provide evidence that cross-cultural differences in similarity
judgments are related to patterns in linguistic statistics (which also
vary across cultures).

@@

There are still many open questions for this account, however. Our
operationalization of environmental statistics with word embeddings
likely captures both language inputs and culture-specific
conceptualization of similarity. Further investigation of the semantic
spaces captured by cross-cultural word embedding models is required.
Future work should also aim to provide a more specific computational
account of how lexical co-occurrence might guide categorization
preference beyond the simple proportion-of-similarity model tested here,
and investigate other person- or triad-specific factors that might
influence how environmental statistics affect similarity judgments.
\textbf{\emph{This work should also investigate person- and
item-specific factors, and characterize responses that are
representative of each group compared to the others.}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-hofstede}{}}%
{Country Comparison - Hofstede Insights}. (2018). \emph{Hofstede
Insights}. Retrieved from
\url{https://www.hofstede-insights.com/country-comparison-tool?countries=china\%2Cunited\%2Bstates\%2Cvietnam}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Hirschman1996}{}}%
Hirschman, C., \& Loi, V. M. (1996). Family and household structure in
vietnam: Some glimpses from a recent survey. \emph{Pacific Affairs},
\emph{69}(2), 229--249. Retrieved from
\url{http://www.jstor.org/stable/2760726}

@@

\leavevmode\vadjust pre{\hypertarget{ref-McDonald2001}{}}%
McDonald, S., \& Ramscar, M. (2001). Testing the distributional
hypothesis: The influence of context on judgements of semantic
similarity. In.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Minkov2012}{}}%
Minkov, M., \& Hofstede, G. (2012). Is national culture a meaningful
concept?: Cultural values delineate homogeneous national clusters of
in-country regions. \emph{Cross-Cultural Research}, \emph{46}(2),
133--159.
http://doi.org/\href{https://doi.org/10.1177/1069397111427262}{10.1177/1069397111427262}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Whitmore1984}{}}%
Whitmore, J. K. (1984). Social organization and confucian thought in
vietnam. \emph{Journal of Southeast Asian Studies}, \emph{15}(2),
296--306. Retrieved from \url{http://www.jstor.org/stable/20070597}

\leavevmode\vadjust pre{\hypertarget{ref-Woodside1998}{}}%
Woodside, A. (1998). Territorial order and collective-identity tensions
in confucian asia: China, vietnam, korea. \emph{Daedalus},
\emph{127}(3), 191--220. Retrieved from
\url{http://www.jstor.org/stable/20027512}
